On Optimality of Finite Element Integration

### 2015-09-22 TOMS_Paper/assembly.tex
Start rewriting section 3...

Pre-evaluation opportunities could have to be exposed, for instance through an exploration of the expression tree transformation space, which can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities
\end{enumerate}

We tackle these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2, it is useful to conclude this section refining our optimality statement as follows.

\begin{Def}[Optimality of a loop nest with bounded working set]
\label{def:optimality}
The synthesis of a loop nest is optimal if, under a set of memory constraints $C$, the number of operations performed in the loop nest is minimal.
\end{Def}

\section{Optimal Integration Routines}
\label{sec:optimal-synthesis}

In this section, we will reason at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.

Our point of departure is the example loop nest in Figure~\ref{code:loopnest}. We make the following observations. 1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.

% maybe say somewhere the order of the loop nest eijk is just the most natural --- the other present worse data locality for example

\subsection{Memory constraints}
\label{sec:mem-const}
The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $\Lambda$. Imagine $\Lambda$ is enclosed in a time stepping loop. One could think of exposing and then hoisting time-invariant sub-expressions. For those sub-expressions that depend on the mesh geometry (i.e., they would depend on $L_e$), code motion would increase the working set by a factor $L$. The gain in number of operations executed could then be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.

A second, more general fact is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $\Lambda$, requires temporary arrays of size $N\cdot O$. This can sometimes break the local memory threshold. We will clarify and experiment this in Section~\ref{sec:perf-results-forms}.

Based on these considerations, we characterize the set of memory constraints $C$ (see Definition~\ref{def:optimality}) as follows
\begin{enumerate}
\item The size of a temporary due to code motion cannot be larger than that of the multilinear iteration space. 
\item The total amount of memory occupied by the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
\end{enumerate}
A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which we discussed at the beginning of this section, is now forbidden. Consequently, the search space for optimality becomes smaller.

\subsection{Minimizing the Operation Count in Inner Loops}
Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation counts is minimum. We now discuss how we can systematically achieve this.

Eliminating sharing from the inner multilinear loops does not suffice. In fact, as suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries. As part of this pre-evaluation, the reduction induced by $L_i$ vanishes. In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.

The challenge is to understand when, and for which monomials, pre-evaluation is profitable. We propose an algorithm and a discussion of its optimality. The intuition of the main algorithm for reducing a loop nest to optimal form is shown in Figure~\ref{code:intuition}.

\begin{figure}[h]
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the main algorithm}
\label{code:intuition}
\end{figure}

@@

\subsection{A-Priori Operation Counting}
\label{sec:op_count}
It remains to tie one loose hand: the construction of the pre-evaluation cost function $\theta$. We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the sharing-free operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned. Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.

>>>>

Pre-evaluation opportunities could be exposed through an exploration of the expression tree transformation space. For arbitrary loop nests, this can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could prevent pre-evaluation; pre-evaluation might expose further sharing)
\end{enumerate}

We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude this section refining our optimality statement as follows.

\begin{Def}[Optimality of a loop nest with bounded working set]
\label{def:optimality}
Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and a set of memory constraints is satisfied. We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations that it performs to evaluate the result is minimal.
\end{Def}

\section{Transformation Algorithm}
\label{sec:optimal-synthesis}

In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form. A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation, the transformation introduced in the previous section (and later further expanded). The optimality of the transformation algorithm is discussed.

\subsection{Memory constraints}
\label{sec:mem-const}
We provide intuitions about the need for memory constraints. Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}. Analogous considerations apply to forms of arbitrary arity.

The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions. Imagine $\Lambda$ is enclosed in a time stepping loop. One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step. The working set size could then increase by a factor $L$. The gain in number of operations executed could therefore be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break a ``local memory threshold''. We will report results of our experiments on the memory threshold in Section~\ref{sec:perf-results-forms}.

Based on these considerations, we establish the following set of memory constraints
\begin{enumerate}
\item The size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space. 
\item The total amount of memory occupied by the temporaries created by code motion cannot exceed a threshold \texttt{$T_H$}
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

\subsection{Minimization of the Operation Count}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

We reinforce that eliminating sharing from the outer-product loop nest does not suffice. As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries. As part of this process, the reduction induced by $L_i$ is computed and, therefore, disappears from $\Lambda$. This is actually what we were referring to when we introduced the pre-evaluation transformation in Section~\ref{sec:optimal-impl}.

There are two issues: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter prevents the former. We start with the second issue. For this, it is convenient to introduce the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion on its optimality follow.

\begin{figure}[h]
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the transformation algorithm}
\label{code:intuition}
\end{figure}

@@

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie one loose hand: the construction of the pre-evaluation cost function $\theta$. We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the sharing-free operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned. Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
